---
title: "RunYourAnalysis_MouseToHuman"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Setup

```{r message=FALSE, warning=FALSE}
# The trimmed required.R loads only the packages the notebook needs
# (here, tidyverse, janitor, tidymodels, clusterProfiler, plotly).
#Install clusterProfiler (once)
library(here)
source(here("notebook", "required.R"))
```

# Load the trained models

The models were trained on mouse-to-human transfer across infection and
injury conditions. Here we apply them to a new (or the same) mouse
experiment.

```{r}
# Model choice: the Lasso is the best-performing model for shared-LEG
# classification in the companion atlas (ROC-AUC 0.617 predictive / 0.627
# explanatory, vs 0.572/0.622 for the random forest), so it is used here for
# score_shared. Rank and direction use the random forest (best for those tasks).
model_shared    <- readRDS(here("notebook", "lasso_model_shared.rds"))
model_rank      <- readRDS(here("notebook", "rf_model_rank.rds"))
model_direction <- readRDS(here("notebook", "rf_model_direction.rds"))
```

# Prepare features table

Please ensure that your input is a table containing the following
columns: pathogen, treatment, timepoint, hgnc_symbol, mean_l2fc,
ci_lower, ci_upper, ave_expr, t, p_value, adj_p_val, b, se, sd,
inverse_se, organism, and condition.

```{r}
# Gene expression data (mouse experiment; example: S. aureus, timepoint 12)
dge_input <- readRDS(here("notebook", "example_dge_result.rds"))

# Protein identity
alignments_human2mouse_all <- readRDS(file = here("notebook", "alignments_human2mouse_all.rds"))

# Distance
human_mouse_cds_distance <- readRDS(file = here("notebook", "human_mouse_cds_distance.rds"))

# Transcription factors
gene_tf_shared_unique <- readRDS(file = here("notebook", "gene_tf_shared_unique.rds"))

# CREs
cres_type_homology_comparison_wide <- readRDS(file = here("notebook", "cres_type_homology_comparison_wide.rds"))

# BTMs annotation
btm_annotation_genes <- read_csv(here("notebook", "btm_annotation_genes.csv")) %>%
  select(hgnc_symbol = symbol, process, group, subgroup) %>%
  distinct()
```

# Build the feature table

```{r}
# Non-BTM genes (kept as a category)
btm_annotation_genes_others <- bind_rows(
  btm_annotation_genes,
  dge_input %>%
    select(hgnc_symbol) %>%
    anti_join(btm_annotation_genes %>%
                select(hgnc_symbol) %>%
                distinct(),
              by = "hgnc_symbol") %>%
    mutate(process = "Not-BTM",
           group = "Not-BTM",
           subgroup = "Not-BTM")
)

# BTM annotation in wide format (one column per module group)
btm_annotation_genes_wide <- btm_annotation_genes_others %>%
  select(hgnc_symbol, group) %>%
  group_by(hgnc_symbol, group) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = group,
              values_from = n) %>%
  column_to_rownames("hgnc_symbol") %>%
  as.matrix() %>%
  replace(is.na(.), 0) %>%
  as.data.frame() %>%
  rownames_to_column("hgnc_symbol")

# Merge all biological feature layers (wide format)
human_mouse_gene_annotation_wide <- human_mouse_cds_distance %>%
  select(hgnc_symbol, dist_k80) %>%
  inner_join(alignments_human2mouse_all,
             by = join_by("hgnc_symbol" == "gene")) %>%
  inner_join(btm_annotation_genes_wide,
             by = "hgnc_symbol") %>%
  inner_join(gene_tf_shared_unique,
             by = join_by("hgnc_symbol" == "gene")) %>%
  inner_join(cres_type_homology_comparison_wide %>%
               select(hgnc_symbol, 'n_total_cres_gene':'pct_match_ctcf_PLS_CTCF-bound'),
             by = join_by(hgnc_symbol))
```

# GSEA (leading-edge genes)

Rank genes with a composite metric (log2FC x -log(adj_p_val)) and run
GSEA against the BTM gene sets.

```{r message=FALSE, warning=FALSE}
# Prepare BTM term-to-gene mapping for GSEA
term2gene <- btm_annotation_genes %>%
  select(process, genes = hgnc_symbol)

# Build ranked gene list with composite metric
df_genes2GSEA <- dge_input %>%
  distinct() %>%
  select(condition,
         genes = hgnc_symbol,
         log2fold_change = mean_l2fc,
         adj_p_val) %>%
  filter(!is.na(genes), !is.na(log2fold_change)) %>%
  mutate(log2fold_change_logp = log2fold_change * -log(adj_p_val)) %>%
  select(condition, genes, rank = log2fold_change_logp)

# Run GSEA across all conditions
gsea_output <- autoGSEA(df = df_genes2GSEA,
                        TERM2GENE = term2gene,
                        geneset_name = "BTMs")

# Clean GSEA output and merge with experiment metadata
gsea_results <- gsea_output$gsea %>%
  as.data.frame() %>%
  clean_names() %>%
  rename(process = id) %>%
  filter(!is.na(process)) %>%
  rownames_to_column("del") %>%
  select(-del, -description) %>%
  rename(p_adjust_gsea = p_adjust,
         q_value_gsea = qvalue,
         rank_gsea = rank) %>%
  inner_join(dge_input %>%
               select(pathogen, treatment, timepoint, condition, organism) %>%
               distinct(),
             by = "condition")

gsea_results
```

```{r}
# Get leading-edge genes (LEGs)
legs <- gsea_results %>%
  separate_rows(core_enrichment, sep = "/") %>%
  mutate(hgnc_symbol = core_enrichment,
         is_leg = "LEG") %>%
  inner_join(dge_input,
             by = join_by(organism, pathogen, treatment, timepoint, condition, hgnc_symbol))

legs %>%
  dplyr::count(process, is_leg)
```

# Compute the model features

The models expect mouse-derived features. Here we rename the columns and
compute the absolute rank, the BTM-only rank and the direction sign.

```{r}
model_input <- dge_input %>%
  # Rename to the names expected by the models
  rename(mean_log2fc_mouse = mean_l2fc,
         inverse_se_mouse = inverse_se,
         p_adj_mouse = adj_p_val,
         timepoint_comparison = timepoint) %>%
  # Direction of the mouse change
  mutate(sign_mouse = sign(mean_log2fc_mouse)) %>%
  # Absolute rank: percentile of |log2FC| x -log10(adjusted P), per condition
  group_by(treatment, pathogen, timepoint_comparison) %>%
  mutate(rank_mouse = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100) %>%
  ungroup() %>%
  # Annotate the biological layers
  inner_join(human_mouse_gene_annotation_wide,
             by = "hgnc_symbol")

# BTM-only absolute rank (rank among genes that belong to a BTM)
btm_gene_set <- btm_annotation_genes %>%
  distinct(hgnc_symbol) %>%
  pull(hgnc_symbol)

model_input <- model_input %>%
  filter(hgnc_symbol %in% btm_gene_set) %>%
  group_by(treatment, pathogen, timepoint_comparison) %>%
  mutate(rank_mouse_btm = percent_rank(abs(mean_log2fc_mouse) * -log10(pmax(p_adj_mouse, 1e-300))) * 100) %>%
  ungroup()

model_input
```

# Predict human rank, direction and shared status

```{r}
# Predictions from the three models (probability for classification, value for regression)
predictions <- model_input %>%
  bind_cols(
    predict(model_rank, new_data = model_input, type = "numeric") %>% rename(score_rank = .pred)
  ) %>%
  bind_cols(
    predict(model_direction, new_data = model_input, type = "prob") %>% select(score_direction = .pred_TRUE)
  ) %>%
  bind_cols(
    predict(model_shared, new_data = model_input, type = "prob") %>% select(score_shared = .pred_TRUE)
  ) %>%
  mutate(score_translational = score_rank * score_direction)

predictions %>%
  select(hgnc_symbol, mean_log2fc_mouse, p_adj_mouse, rank_mouse, inverse_se_mouse, rank_mouse_btm, sign_mouse,
         score_shared, score_rank, score_direction, score_translational) %>%
  arrange(desc(score_translational))
```

# Heatmap helpers

```{r}
# Features shown in the heatmaps (readable label = column name)
heat_map <- c(
  "Log2FC (mouse)"        = "mean_log2fc_mouse",
  "Absolute rank (mouse)" = "rank_mouse",
  "1/SE"                  = "inverse_se_mouse",
  "Protein identity"      = "identity_human2mouse",
  "Kimura"                = "dist_k80",
  "nTotal TFs"            = "n_tf_total",
  "%Shared TFs"           = "pct_tf_shared",
  "nTotal cCREs"          = "n_total_cres_gene",
  "%Match Type (PLS)"     = "pct_match_type_PLS",
  "%Match Type (pELS)"    = "pct_match_type_pELS",
  "%Match Type (dELS)"    = "pct_match_type_dELS",
  "CTCF match"            = "pct_match_ctcf_dELS_CTCF-bound"
)

features_order<- c(
  "Log2FC (mouse)",
  "Absolute rank (mouse)",
  "1/SE",
  "Protein identity",
  "Kimura",
  "nTotal TFs",
  "%Shared TFs",
  "nTotal cCREs",
  "%Match Type (PLS)" ,
  "%Match Type (pELS)",
  "%Match Type (dELS)",
  "CTCF match"
)

build_heat_mat <- function(df) {
  df %>%
    distinct(hgnc_symbol, .keep_all = TRUE) %>%
    select(hgnc_symbol, all_of(unname(heat_map))) %>%
    setNames(c("hgnc_symbol", names(heat_map))) %>%
    column_to_rownames("hgnc_symbol") %>%
    as.matrix()
}

# Rows = genes, columns = features. z-scores are computed relative to `baseline_mat`
# (all genes) so the colours are comparable across different gene selections.
feature_heatmap <- function(genes_df, baseline_mat) {
  base_mean <- colMeans(baseline_mat, na.rm = TRUE)
  base_sd   <- apply(baseline_mat, 2, sd, na.rm = TRUE)
  heat_dat  <- build_heat_mat(genes_df)
  heat_z    <- sweep(sweep(heat_dat, 2, base_mean, "-"), 2, base_sd, "/")
  heat_z[is.na(heat_z)] <- 0
  gene_levels <- rownames(heat_z)[hclust(dist(heat_z))$order]
  feat_levels <- colnames(heat_z)[hclust(dist(t(heat_z)))$order]
  heat_dat %>%
    as.data.frame() %>%
    rownames_to_column("hgnc_symbol") %>%
    pivot_longer(-hgnc_symbol, names_to = "feature", values_to = "value") %>%
    mutate(z = (value - base_mean[feature]) / base_sd[feature],
           hgnc_symbol = factor(hgnc_symbol, levels = gene_levels),
           feature = factor(feature, levels = feat_levels))
}

# Score heatmap: rows = genes, columns = the four model scores (z-scored over ALL genes).
score_heat_cols <- c(
  "Shared"        = "score_shared",
  "Rank"          = "score_rank",
  "Direction"     = "score_direction",
  "Translational" = "score_translational"
)

score_heatmap <- function(genes_df, baseline_df, title) {
  score_mat <- function(df) {
    df %>%
      distinct(hgnc_symbol, .keep_all = TRUE) %>%
      select(hgnc_symbol, all_of(unname(score_heat_cols))) %>%
      setNames(c("hgnc_symbol", names(score_heat_cols))) %>%
      column_to_rownames("hgnc_symbol") %>%
      as.matrix()
  }
  base_mat  <- score_mat(baseline_df)
  base_mean <- colMeans(base_mat, na.rm = TRUE)
  base_sd   <- apply(base_mat, 2, sd, na.rm = TRUE)
  heat_dat  <- score_mat(genes_df)
  heat_z    <- sweep(sweep(heat_dat, 2, base_mean, "-"), 2, base_sd, "/")
  heat_z[is.na(heat_z)] <- 0
  gene_levels <- rownames(heat_z)[hclust(dist(heat_z))$order]
  heat_dat %>%
    as.data.frame() %>%
    rownames_to_column("hgnc_symbol") %>%
    pivot_longer(-hgnc_symbol, names_to = "score", values_to = "value") %>%
    mutate(z = (value - base_mean[score]) / base_sd[score],
           hgnc_symbol = factor(hgnc_symbol, levels = gene_levels),
           score = factor(score, levels = colnames(heat_dat)))
}
```

# Visualise the scores

```{r fig.height=4, fig.width=12}
# score_rank: predicted human absolute rank
score_rank_plot = predictions %>%
  ggplot() +
  aes(x = score_rank, y = rank_mouse, text = hgnc_symbol) +
  geom_point(shape = 21, size = 2,  
             alpha = 0.5, stroke = 0.2,
             fill = "#4DBBD5FF",
             color = "black") +
  theme_vaxgo() +
  theme(legend.position = "top") +
  labs(x = "Score (Rank)",
       y = "Absolute rank (mouse)",
       title = "Rank score")

# score_direction: probability of concordant direction (same sign in both species)
score_direction_plot =
predictions %>%
  ggplot() +
  aes(x = score_direction, y = rank_mouse, text = hgnc_symbol) +
  geom_point(shape = 21, size = 2,  
             alpha = 0.5, stroke = 0.2,
             fill = "#4DBBD5FF",
             color = "black") +
  theme_vaxgo() +
  theme(legend.position = "top") +
  labs(x = "Score (Direction)",
       y = "Absolute rank (mouse)",
       title = "Direction score")

# score_translational: combined score (high rank AND concordant direction)
score_translational_plot = predictions %>%
  ggplot() +
  aes(x = score_translational, y = rank_mouse, text = hgnc_symbol) +
  geom_point(shape = 21, size = 2,  
             alpha = 0.5, stroke = 0.2,
             fill = "#4DBBD5FF",
             color = "black") +
  theme_vaxgo() +
  theme(legend.position = "top") +
  labs(x = "Score (Translational)",
       y = "Absolute rank (mouse)",
       title = "Translational score")

#Static visualization
score_rank_plot + score_direction_plot + score_translational_plot +
  plot_layout(axes = "collect") +
  plot_annotation(theme = theme(legend.position = "top"))
```

Make it interactive
```{r fig.height=3, fig.width=3}
#Make it interactive
score_rank_plot %>% ggplotly(tooltip = c("text", "x", "y", "fill"))
score_direction_plot %>%ggplotly(tooltip = c("text", "x", "y", "fill"))
score_translational_plot %>% ggplotly(tooltip = c("text", "x", "y", "fill"))
```

# Select priority genes with a user-defined cutoff

```{r}
# Example: keep genes above a cutoff you choose
score_cutoff <- 0.8

priority_genes <- predictions %>%
  filter(score_translational >= score_cutoff) %>%
  arrange(desc(score_translational)) %>%
  select(hgnc_symbol, score_shared, score_rank, score_direction, score_translational)

priority_genes
```

## Gene heatmaps

```{r fig.height=5, fig.width=12}

# Genes shown in the gene-level heatmaps: top-N by translational score.
# (The score table has one row per gene and condition, so we keep one row per gene.)
n_heat_genes <- 25

heat_genes <- predictions %>%
  distinct(hgnc_symbol, .keep_all = TRUE) %>%
  slice_max(score_translational, n = n_heat_genes)

# Rows = genes, columns = the four model scores (z-scored over ALL genes)

selected_genes_scores_plots = heat_genes %>% 
    ggplot() +
    aes(x = score_direction,
        y = score_rank, 
        fill = score_translational,
        size = score_translational,
        label = hgnc_symbol) +
    geom_point( color = "black", pch = 21) +
    scale_fill_gradientn(colours = c("white", "#4361ee")) +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
          axis.text.y = element_text(size = 10),
          legend.position = "top") +
    labs(x = "Score Direction", 
         y = "Score Rank", 
         fill = "Score Translational",
         size = "Score Translational") +
    geom_text_repel(
    aes(label = hgnc_symbol),
    size = 3,
    color = "black",              
    fill = alpha("white", 0.6),    
    label.r = unit(0.15, "lines"), 
    max.overlaps = 100,  
    direction = "both",    
    # hjust = 0,                 
    box.padding = 0.5, 
    point.padding = 0.3
  )+
  scale_x_continuous(limits = c(NA, 1.01)) 

selected_genes_scores_plots %>% ggplotly()

# Rows = genes, columns = the biological features (z-scored over ALL genes)

selected_genes_features_heatmap = feature_heatmap(heat_genes, build_heat_mat(predictions)) %>%
  mutate(feature = fct_relevel(feature, features_order) %>% fct_rev()) %>%  
    ggplot() +
    aes(x =hgnc_symbol , y = feature, fill = z,
        text = paste0(hgnc_symbol, "\n", feature, ": ", round(value, 2))) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_gradientn(colours = c("#4DBBD5FF", "white", "#4361ee"),
                         limits = c(-2.5, 2.5), oob = scales::squish) +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
          axis.text.y = element_text(size = 10),
          legend.position = "top") +
    labs(x = "", y = "", 
         fill = "Z-score, all genes", 
         title = "Selected genes (features)")

selected_genes_features_heatmap +
selected_genes_scores_plots +
  plot_annotation(theme = theme(legend.position = "right"),
                  title = "Selected genes")

```

# Heatmaps

## Enriched processes (NES)

```{r fig.height=20, fig.width=5}
# padj cutoff used to keep processes in the heatmap
padj_filter <- 0.25

process_nes <- gsea_results %>%
  filter(p_adjust_gsea < padj_filter, !is.na(nes))

if (nrow(process_nes) > 0) {
  process_heatmap <- process_nes %>%
    mutate(process = fct_reorder(process, nes)) %>%
    ggplot() +
    aes(x = condition, y = process, fill = nes) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_gradient2(low = "#4361ee", mid = "white", high = "#e5383b", midpoint = 0) +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
          axis.text.y = element_text(size = 7)) +
    labs(x = "", y = "", fill = "NES",
         title = paste0("Enriched processes (padj < ", padj_filter, ")"))
  process_heatmap
} else {
  message("No enriched processes pass padj < ", padj_filter, ".")
}
```

# Selected process

Set `selected_process` to one of the enriched process names printed
below, then re-knit the document.

```{r}
# Enriched processes available to choose from (copy one into `selected_process`)
gsea_results %>%
  filter(p_adjust_gsea < padj_filter, !is.na(nes)) %>%
  distinct(process) %>%
  arrange(process) %>%
  pull(process)
```

```{r fig.height=6, fig.width=12}
# --- Choose a process --------------------------------------------------------
selected_process <- "T cell activation and signaling"

# Leading-edge genes (LEGs) of the selected process
process_genes <- gsea_results %>%
  filter(process == selected_process) %>%
  pull(core_enrichment) %>%
  strsplit("/") %>%
  unlist() %>%
  unique()

# --- Shared score vs mouse rank for the LEGs of the selected process ---------
process_scores <- predictions %>%
  filter(hgnc_symbol %in% process_genes)

# --- Feature heatmap of the LEGs of the selected process ---------------------
# z-scores are computed within the process/module genes (not across all genes)
process_scores_mat = process_scores %>%
    distinct(hgnc_symbol, .keep_all = TRUE) %>%
    select(hgnc_symbol,all_of(unname(heat_map))) %>%
    setNames(c("hgnc_symbol", names(heat_map))) %>%
    column_to_rownames("hgnc_symbol") %>%
    as.matrix() 

process_features_heatmap = feature_heatmap(process_scores, process_scores_mat) %>% 
  ggplot() +
    aes(x = feature, y = hgnc_symbol, fill = z,
        text = paste0(hgnc_symbol, "\n", feature, ": ", round(value, 2))) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_gradientn(colours = c("#4DBBD5FF", "white", "#4361ee"),
                         limits = c(-2.5, 2.5), oob = scales::squish) +
    theme_vaxgo() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
          axis.text.y = element_text(size = 10),
          legend.position = "top") +
    labs(x = "", y = "", 
         fill = "Z-score, process genes", 
         title = paste0("Process: ", selected_process))


process_score_plot <- process_scores %>%
    ggplot() +
    aes(x = score_shared, y = rank_mouse, fill = score_shared,
        text = paste0(hgnc_symbol,
                      "\nscore_shared: ", round(score_shared, 2),
                      "\nrank_mouse: ", round(rank_mouse, 1))) +
    geom_point(shape = 21, size = 2, colour = "black", stroke = 0.3, alpha = 0.85) +
  geom_text_repel(aes(label = hgnc_symbol)) +
    scale_fill_gradient(low = "white", high = "#4DBBD5FF") +
    theme_vaxgo() +
    theme(legend.position = "right") +
    labs(x = "Score (Shared)", 
         y = "Absolute rank (mouse)",
         fill = "Score (Shared)",
         title = paste0("Shared score - ", selected_process))



#Plot all
process_features_heatmap|process_score_plot 


#Interactive
process_score_plot %>% 
ggplotly(tooltip = "text") %>%
    layout(margin = list(l = 60, r = 15, b = 55, t = 40))

process_features_heatmap %>% 
  ggplotly(tooltip = "text") %>%
    layout(margin = list(l = 60, r = 15, b = 55, t = 40))

```
